Sperm should evolve to make female meiosis fair.

### 2014-07-16 writeup_and_notes/GeneticsResub/male_dependent_drive.tex
graham's changes to intro

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations
	likely precludes this interaction in plants). 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for exploitation by such selfish systems, \yb{and here we ask if spam do indeed evolve to collaborate with female drivers to exploit this apparent weakness in the defense against meiotic drive.} \newline 
%Why then is the requirement of fertilization to complete female meiosis so ubiquitous?

\yb{The threat to meiotic fairness posed by the fertilization requirement for completion of female meiosis seems both real and avoidable. 
We say this threat is avoidable because the stage at which female meiosis requires fertilization 
	varies widely across taxa, and in a few animals female meiosis is complete upon ovulation \
	(that is, fertilization is not required for female meiosis \citep[e.g. Table 1 in ][]{Masui_book}). 
We say this threat is real because there is substantial evidence that female meiotic drive is an 
	evolutionary force and considerable opportunity for sperm to influence female drive.  
While research to date has identified a few of what are likely the most extreme female meiotic 
	drivers \citep{Agulnik1990, Fishman2008, Hiatt2003,Novitski1951, Pardo-ManuelDeVillena2001c}, 
	rapid evolution of centromeres, telomeres and other basic components of the meiotic machinery, 
	suggest consistent evolution of meiotic drivers and suppressors of meiotic drive
	 \citep[e.g. ][]{Anderson2008, Anderson2009, Axelsson2010, Malik2009a}, and the inexpense of 
	 genotyping-by-sequencing platforms for non-model systems will likely 
	identify subtle meiotic drivers in numerous non-model systems.  
Both mechanistic and genetic evidence suggests that sperm can influence the outcome of female meiosis.
An example of mechanistic evidence for sperm's influence on female meiosis comes from \emph{C. elegans}, where premature deployment of the aster -- a vital 	component of mitotic machinery provided by the sperm, 
	disrupts MII meiotic segregation in the egg, leading to a triploid zygote \citep{McNally2012}.
{\bf{GRAHAM DO MORE????} } 
Genetic evidence from the \emph{In} and \emph{Om}  drive systems in mice suggests that the 
	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages
	52-54 in][]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the 	
	genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}. 
}\newline

@@

We develop population genetic models exploring the evolution of sperm influence on female meiotic drive. 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such models universally reveal that sperm-acting drive modifiers are favored only if they suppress drive. 
These results highlight the fact that the interests of sperm and maternal  genomes' are often aligned, as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Thus, there is little selective benefit to females in preventing sperm to influence female meioses. 
\yb{While we do not claim that female meiotic arrest evolved specifically to recruit helpful sperm to defend against meiotic drive, the delayed completion of female meiotic drive will not in fact encourage sperm to assist renegade drivers.}  
%Rather, features of o\"{o}genesis  may evolve to facilitate an interaction with sperm %with female meiosis
%	as yet another mechanism to enforce an equitable transfer of genetic material.

>>>>

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, female meiosis is completed only after fertilization \citep{Masui_book}, 
	creating ample opportunity for interaction between the sperm and
	female meiotic machinery (note that the alternation of generations
	likely precludes this interaction in plants). 
\gc{It is becoming
        increasingly clear that sperm bring a wide variety of RNA and proteins into the
        egg REF. Some of these have known functions, for example, in most
        animal species sperm-- not eggs-- are
        responsible for the transmission of the centrosome, a vitial
        component of the mitotic machinary for the zygote. However,
        the function of many of th products the sperm beings into the
        egg is completely unknown and 
      these products vary widely over species. 
The existance of paternal effect mutations in model systems highlights
 the fact that some of these components have a wide-range of functions in
 the completion of synagamy and the development of the zygote. 
Thus sperm could potentially have functional effects on
various aspects of female meiosis, e.g. in some systems 
sperm actually carry the factor responsibly for the resumption of
female meiosis, rather than this being a simple consequence of egg
activation. } 
Therefore, a `green-bearded' \citep{Gardner2010} sperm-acting allele that recognizes and facilitates the meiotic drive of a genetically equivalent allele in heterozygous females %female meiosis %(a  heteromorphic dyad), 
     could presumably rapidly spread through a population. 
At first sight, female meiosis appears primed for exploitation by such selfish systems. Here we ask if sperm do indeed evolve to collaborate with female drivers to exploit this apparent weakness in the defense against meiotic drive. \newline 
%Why then is the requirement of fertilization to complete female meiosis so ubiquitous?

It is not the case that sperm is mechanistically incapable of influencing female meiosis.
Both mechanistic and genetic evidence suggests that sperm can influence the outcome of female meiosis.
An example of mechanistic evidence for sperm's influence on female meiosis comes from \emph{C. elegans}, where premature deployment of the aster -- a vital 	component of mitotic machinery provided by the sperm, 
	disrupts MII meiotic segregation in the egg, leading to a triploid zygote \citep{McNally2012}.
There are current only two polymorphic,characterized females meiotic drive systems in
animals-- and both of them, the \emph{In} and \emph{Om}  drive systems in mice-- suggests that the 
	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages
	52-54 in][ for comment]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the 	
	genotype of the fertilizing sperm
        \citep{Agulnik1993,Wu2005}. 
The fact that both of polymorphic, putative female meiotic drive systems identified
to date in animals (\emph{In} and \emph{Om} in mice) seem to show an
effect of sperm genotype suggests that if female meiotic drive is
common the role of sperm will be important as well.

There is good evidence that female meiotic drive is an 
	evolutionary force and offering considerable opportunities for sperm to influence female drive.  
While research to date has identified a few extreme female meiotic 
	drivers in the small number model systems systematically
        studied to date \citep{Agulnik1990, Fishman2008, Hiatt2003,Novitski1951, Pardo-ManuelDeVillena2001c}, 
	rapid evolution of centromeres, telomeres and other basic components of the meiotic machinery, 
	suggest consistent evolution of female meiotic drivers and suppressors of meiotic drive
	 \citep[e.g. ][]{Anderson2008, Anderson2009, Axelsson2010,
           Malik2009a}. We expect that over the next decade the spread of sequencing to a
         range of systems will reveal many more female meiotic drive
         systems, however, carefully characterizing them will still
         remain a challenging task.
\newline

@@

We develop population genetic models exploring the evolution of sperm influence on female meiotic drive. 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such models universally reveal that sperm-acting drive modifiers are favored only if they suppress drive. 
These results highlight the fact that the interests of sperm and
maternal  genomes' are often aligned, 
as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Thus, there is little selective benefit to females in preventing sperm to influence female meioses. 
\gc{We do not claim that female meiotic arrest evolved
  specifically to recruit helpful sperm to defend against meiotic
  drive, the delayed completion of female meiotic drive will not in
  fact encourage sperm to assist renegade drivers. Rather, we think
  that features of
  o\"{o}genesis  may evolve to facilitate an interaction with sperm 
with female meiosis
as yet another mechanism to enforce an equitable transfer of genetic material.}
